Analysis of finite-volume advection schemes on cubed-sphere grids and an accurate \textcolor{red}{scheme} for divergent winds

### 2024-05-21 paper.tex
updated intro

%% Abstract
\begin{abstract}
%% Text of abstract
To be written.
\end{abstract}

@@

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{intro}
The Finite Volume Cubed-Sphere Dynamical Core (FV3), developed by the Geophysical Fluid Dynamics Laboratory (GFDL) and the National Oceanic and Atmospheric Administration (NOAA), 
has been embraced as the dynamical core for several atmospheric models (cf. eg. \cite{zhou:2015,lee:2020, bertrand:2020,harris:2020,martin:2022,zhang:2024}).
In 2019, FV3 attained significant recognition when it was adopted as the new dynamical core for the Global Forecast System the U.S. National Weather Service (NWS), replacing the spectral transform GFS.
Additionally, FV3 is used in the Hurricane Analysis and Forecast System (HAFS) from the NWS for tropical cyclone forecasts \cite{dong:2020}.
 % (\url{https://www.noaa.gov/media-release/noaa-upgrades-us-global-weather-forecast-model}, last
%accessed on April 19th, 2024).

@@

On the other hand, within FV3, the solution to the SWEs is derived using the method proposed by \cite{lin:1997} extended to the cubed-sphere grid.
This approach takes into account the SWEs in their vector invariant form, along with the C and D-grids introduced by \cite{arakawa:1977}.
A major feature of this scheme is that the computation of fluid pressure, absolute vorticity, and kinetic energy fluxes is carried out using only transport finite-volume fluxes. 
Thus, the transport scheme assumes a critical role in shaping the horizontal dynamics of FV3, beyond its traditional function in the dynamical core, as observed in tracer transport \cite{will:2007}.

@@

\textcolor{red}{
Therefore, given the relevance of the FV3 dynamical core,
the goal of this study is to reassess and suggest enhancements for the current FV3 transport scheme originally developed by \cite{putman:2007}, given its pivotal role in the horizontal dynamics of FV3.}
It is demonstrated in this work that the scheme proposed by \cite{putman:2007} assumes constant metric terms during the application of the PPM to each 1D flux integration domain and employs a first-order departure point calculation for the 1D fluxes.
\textcolor{red}{
The development of a scheme that incorporates a second-order departure point calculation for the 1D fluxes and eliminates the assumption of constant metric terms is proposed.}
While the proposed scheme does not retain the property of splitting error elimination for divergence-free winds and constant scalar fields, it introduces only second-order errors in such scenarios.

>>>>

%% Abstract
\begin{abstract}
%% Text of abstract
The finite-volume dynamical core developed by GFDL-NOAA-USA, originally designed for latitude-
longitude grids, was adapted to the cubed-sphere to enhance scalability on massively parallel supercomputers,
resulting in the creation of the FV3 dynamical core. FV3 serves as the dynamical core for many models
worldwide, and in 2019, it was officially designated as the dynamical core for the new Global Forecast System
of the National Weather Service in the USA, replacing the spectral model. The finite-volume approach
employed by FV3 to solve horizontal dynamics involves applying advection fluxes for different variables;
thus, the advection scheme plays a key role in the model. Therefore, in this thesis, we propose to investigate
the details of the advection scheme of FV3. We were able to suggest modifications to the FV3 advection
scheme that significantly improved advection for divergent winds with only a small extra computational
effort and of simple implementation in the existing codes. We conducted several numerical simulations using
the advection and shallow-water equations. Since the FV3 advection scheme utilizes a splitting strategy,
combining 1D finite-volume flux operators, our improvements were obtained by improving the departure
point computation for the 1D flux operators and modifying the way the metric term of the cubed-sphere is
treated when computing the 1D fluxes. Through simulations, we demonstrate that the current FV3 advection
scheme is only first-order accurate for divergent winds, while our scheme is second-order accurate. For
divergence-free winds, both schemes are second-order, with our scheme being slightly more accurate.
One major difficulty in working on the cubed-sphere is handling coordinate discontinuity along the cube
faces, which may lead to larger errors in these regions. However, we demonstrate through numerical
simulations that the proposed advection scheme exhibits slightly reduced sensitivity to the cube corners. In
conclusion, this thesis offers a comprehensive examination of the FV3 discretization of horizontal dynamics,
providing valuable insights into enhancing the accuracy of the FV3 dynamical core, particularly for divergent
winds.
\end{abstract}

@@

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{intro}
The Finite Volume Cubed-Sphere Dynamical Core (FV3), developed by the Geophysical Fluid Dynamics Laboratory (GFDL) and the National Oceanic and Atmospheric Administration (NOAA), 
has been embraced as the dynamical core for several atmospheric models (cf. eg. \cite{zhou:2015,lee:2020, bertrand:2020,harris:2020,martin:2022,zhang:2024}).
In 2019, FV3 attained significant recognition when it was adopted as the new dynamical core for the Global Forecast System the U.S. National Weather Service (NWS), replacing the spectral transform dynamical core.
Additionally, FV3 is used in the Hurricane Analysis and Forecast System (HAFS) from the NWS for tropical cyclone forecasts \cite{dong:2020}.
 % (\url{https://www.noaa.gov/media-release/noaa-upgrades-us-global-weather-forecast-model}, last
%accessed on April 19th, 2024).

@@

On the other hand, within FV3, the solution to the SWEs is derived using the finite-volume method proposed by \cite{lin:1997} extended to the cubed-sphere grid.
This approach takes into account the SWEs in their vector invariant form, along with the C and D-grid staggering introduced by \cite{arakawa:1977}.
A major feature of this scheme is that the computation of fluid pressure, absolute vorticity, and kinetic energy fluxes is carried out using only transport finite-volume fluxes. 
Thus, the transport scheme assumes a critical role in shaping the horizontal dynamics of FV3, beyond its traditional function in the dynamical core, as usually observed in tracer transport \cite{will:2007}.

@@

\textcolor{red}{
Therefore, given the relevance of the FV3 dynamical core,
the goal of this study is to reassess and suggest enhancements for the current FV3 transport scheme originally developed by \cite{putman:2007}, given its pivotal role in the horizontal dynamics of FV3.}
It is demonstrated in this work that the scheme proposed by \cite{putman:2007} assumes constant metric terms during the application of the PPM to each 1D flux integration domain and employs a first-order departure point calculation for the 1D fluxes.
\textcolor{red}{
A scheme that incorporates a second-order departure point calculation for the 1D fluxes and eliminates the assumption of constant metric terms is proposed.}
While the proposed scheme does not retain the property of splitting error elimination for divergence-free winds and constant scalar fields, it ensures second-order errors in such scenarios.
